\thetitle

### 2020-11-19 writing/introduction.tex
revision of rational
adding landscape description

\section{Does Blame Matter?} \label{sec:introduction}

@@

The paper's second contribution is a set of results from applying the evaluation
method to three distinct blame assignment strategies and approximately
65000 cases (section~\ref{sec:results}): Vitousek et
al.'s strategy of tracking typed/untyped boundaries in Reticulated;
Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
mf-toplas-2009}; and the approach of industrial systems that forego blame
and instead rely on error messages from the
underlying programming language.

>>>>

\section{Does Blame Matter?}

@@

The paper's second contribution is a set of results from applying the evaluation
method to three distinct blame assignment strategies and approximately
72000 cases (section~\ref{sec:results}): Vitousek et
al.'s strategy of tracking typed/untyped boundaries in Reticulated;
Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
mf-toplas-2009}; and the approach of industrial systems that forego blame
and instead rely on error messages from the
underlying programming language.

### 2020-11-19 writing/landscape.tex
revision of rational
adding landscape description

\section{The Landscape of Gradual Typing and Blame}

   By example; adjusted version of the one in OOPSLA sec 2

  \begin{itemize}

  \item Why Blame Assignments Differ.

  \item How Blame Assignments Differ.

  \item A space of GT system design. In terms of error reporting, we identify two axes of GT system
        design: where checking occurs and whether blame is tracked.
    \end{itemize}

>>>>

\section{How to Bring Natural, Transient and Erasure Under the Same Roof}

%% -----------------------------------------------------------------------------
\def\fname#1#2{$\underline{\hbox to #2in{#1}}$\\[-2ex]}

\begin{figure*}[hbt]\footnotesize
  \vspace{-1mm}

\begin{minipage}[t]{2.2in}
\begin{alltt}\fname{untyped/pack-lib}{1.8}
(provide pack _ _ _)

_ _ _  dependencies     _ _ _
_ _ _  and definitions  _ _ _

(define (pack d)
  ;; process JSON data and 
  ;; package it in a dictionary
  ;; (association list)
  (make-hash _ _ _) ;; BUG!)

\end{alltt}
\end{minipage}\begin{minipage}[t]{2.2in}
\begin{alltt}\fname{typed/pack-lib}{1.7}
(provide typed-pack)

(require/typed pack-lib
  [pack 
  (-> JSON 
      (Listof 
       (Pairof Symbol String)))])

(define typed-pack pack) 
\end{alltt}
\smallskip
\begin{alltt}\fname{typed/crypto-pack-lib}{1.7}
(provide crypto-pack)

(require/typed pack-lib
 [pack 
  (-> JSON 
      (Listof 
       (Pairof Symbol String)))])

(define (crypto-pack [d : JSON])
  (pack (encrypt d _ _ _)))

 \end{alltt}
\end{minipage}\begin{minipage}[t]{2.2in}
\begin{alltt}\fname{untyped/client}{1.65}
(require json 
         typed/pack-lib
           typed/crypto-pack-lib)

_ _ _  other dependencies  _ _ _
_ _ _  and definitions     _ _ _

;; read data from files, pack 
;; and share securely

(define public-data 
  (typed-pack 
   (read-json 
    "public-records")))

(define secret-data 
  (crypto-pack 
   (read-json 
    "medical-records")))

 _ _ _ rest of client _ _ _
\end{alltt}\end{minipage}

   \caption{One mix-typed program, three interpretations} \label{fig:example}

  \vspace{-1mm}
\end{figure*}
%% -----------------------------------------------------------------------------

The  program in figure~\ref{fig:example} demonstrates the differences
between natural, transient and erasure using Typed Racket syntax.  It
consists of four modules: the left and right ones are untyped while
the two modules in the middle are typed.

The \texttt{untyped/pack-lib} module (left) provides, among
others, a function \texttt{pack} that consumes JSON data and packages it
in a dictionary. However, by mistake \texttt{pack} returns
a hash table instead of the association list its documentation promises.
The \texttt{client} module (right) uses \texttt{pack} indirectly to
package public and sensitive data it reads from the filesystem. It does so
through two different intermediary typed modules (center). The first,
\texttt{typed/pack-lib}, simply imports \texttt{pack} and immediately
exports as \texttt{typed-pack} asserting that it is a function
that consumes JSON and returns an association list that maps symbols to
strings. The second typed module, \texttt{typed/crypto-pack-lb},
also imports \texttt{pack} and assumes for it the same type as
\texttt{typed/pack-lb} but instead of exporting it directly it wraps in
another function, \texttt{crypto-pack}, that encrypts its input before
passing it to \texttt{pack}.

Under natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
\texttt{pack} is imported in either \texttt{typed/pack-lb} or
\texttt{typed/secure-pack-lb}, the runtime checks that \texttt{pack} is a
function and wraps it in a proxy that enforces its  type.  Analogously
the run time wraps the exports of the two typed modules in proxies that
check that untyped code applies \texttt{typed-pack} and
\texttt{crypto-pack} to arguments of its type. In essence, the
return type checks protect  the typed modules in case they use
\texttt{pack} and, as is the case here, it does not return an association
list. The check of the argument type  treats the
exports of the two typed modules as typed values that need to be protected
when they are used by untyped code.  As this analysis implies, if the
return type check fails then the problem is with
\texttt{untyped/pack-lib}, while if the argument type check fails
responsibility lies with \texttt{client}. The natural semantics keep track
of the responsible party for each check in the corresponding proxy. Thus,
in our example, when the call to \texttt{typed-pack} returns, the return
type check fails and the natural semantics blames \texttt{pack-lib}.

Under the transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
produce an error. In contrast to natural, transient does not rely on
proxies. Instead, it rewrites typed modules to inline checks at appropriate
places. Furthermore, these checks are
partial, that is, they check only the top constructor of a type. In other words
they check types in a  \emph{shallow} manner. In our
example, transient's inlined checks make sure that the imported \texttt{pack} 
is a function and that the result of its call in \texttt{typed/crypto-pack-lib} is a list.   Exactly this
last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
order to pinpoint the responsible party for the failed checked, transient
keeps on the side a map from values to the boundaries between modules they
cross and the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{untyped/pack-lib} to
\texttt{typed/pack-lib} and from \texttt{typed/pack-lib} to
\texttt{client} with the type that appears in the \texttt{required/typed}
forms in the example. Since the failed check corresponds to the return
type of \texttt{pack}, transient blames the sources of the two boundary
crosses, which in this case is \texttt{untyped/pack-lib} for both. In
general though, transient may blame more than one parties. In fact, the
theoretical work of~\citet{gfd-oopsla-2019} shows that there are programs
where transient constructs a blame set that includes modules that are not
relevant to the check failure or that it even misses entirely the responsible
party ---  our experiment aims to examine the pragmatic impact of exactly these
issues on debugging type-level mistakes.

Under erasure semantics, the example does not raise an error due to the
gradual type system; erasure performs no type checks at runtime. Thus the
program may fail with an exception if \texttt{client} tries to use the
result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
fail at all and produce an wrong result that goes undetected by the
gradual type system .  In either case, the erasure semantics provides no
help to the programmer for detecting the bug in our program.

Since, the goal of our experiment is to compare the effectiveness of blame
across the three semantics, it is critical to isolate their
distinguishing characteristics from superficial implementation
differences. Thus we implement all three systems on top of the common
basis of Racket. First, we use the Typed Racket syntax and type
checker for all three semantics. Second, for natural we use Typed Racket as
is while for erasure we run programs with plain Racket after erasing any
type annotations. Third, for transient we implement a variant of Typed
Racket that follows closely the semantics from~\citet{vss-popl-2017}.

The first major ingredient of transient Typed Racket is to determine the
Racket implementation of the shallow checks that map to each type.  For
certain types, such as {\tt String} or {\tt (Vectorof Byte)}, transient
can re-use a predicate from the Racket language.  Other types require custom
predicates; for example the type {\tt Index} of valid array indices allows
a range of positive exact integers.  Many custom predicates run in
constant time, but some are more involved.  One linear-time check is for
{\tt (Listof T)} for any element type {\tt T}, because it must distinguish
proper from improper list-like structures.  The
key point, however, is that no shallow check inspects beyond the top
constructor of each type.

The second major ingredient of transient Typed Racket  is the addition of
a pass to the compiler of Typed Racket that rewrites all typed code.
Rewriting modifies all elimination forms to shallow-check their result.
For example, if the variable {\tt f} has the static type {\tt (-> String
String)} then the application {\tt (f x)} must be rewritten to check for a
string result because {\tt f} could be an untyped function.  Rewriting
also inserts hooks to update the blame map.  When a value crosses a module
boundary, the blame map records a type assumption.  The inlined checks
for elimination forms also extend the blame map with further information. For
instance, the check of the result of {\tt (f x)} adds to the blame map
that the expected type of the result is the return type of the type of
{\tt f}.  If a transient check ever fails, transient Typed Racket walks over 
the information in the blame map and reports a set of blamed components.

\smallskip
\noindent
{\bf Note.} As the common basis for the three semantics, we choose 
Typed Racket's migratory typing. In contrast to the initial formulation of
gradual typing~\cite{st-sfp-2006}, migratory typing does not provide
support for type \texttt{Dyn}, components are either untyped or fully
typed and the gradual type system does not affect the compilation of
untyped components. Despite, these differences, Typed Racekt is the most
mature system for sound mix-typed code
and recent theoretical work has established that
its migratory typing offers a setting for comparing different approaches to gradual
typing~\cite{gf-icfp-2018,gfd-oopsla-2019}.

### 2020-11-19 writing/mutate.tex
revision of rational
mutate: add line in mutator table seperating labels from contents
adding landscape description

\section{How to Make Lots of Mistakes} \label{sec:mutate}

@@

\begin{figure*}
  \begin{tabular}{p{2.1cm} | p{5cm}  | p{5.5cm} }
    {\bf name} & {\bf description} & {\bf example} \\

@@

The first criterion simply validates that the debugging scenario concerns
an actual type-level mistake. However, The second and third criteria require 
more explanation.  The second one stipulates that the mistake is
detectable under Erasure to ensure that it is detectable by all three
semantics.   Whether a scenario produces a run time error largely depends on 
the underlying semantics but \citet{gfd-oopsla-2019}
show that a run time error under Erasure implies a run time
error under both Transient and Natural. Thus the criterion aligns
interesting debugging scenarios with our lowest denominator, Erasure, 
and excludes  from consideration the very interesting scenarios indeed where Natural or
Transient can detect a mistake but where Erasure is powerless. Of course this choice favors
Erasure over Transient and Natural and, for the same reason, Transient over
Natural. We consider some form of bias towards one or the other semantics
unavoidable no matter how we define interesting debugging scenarios and we
opt for tipping the scales in favor of the theoretically weakest
semantics.

@@

With the definition of interesting debugging scenarios in hand, we analyze how effective our mutators
are for turning the benchmarks into interesting debugging scenarios.
Indeed the mutators produce at least 72192 interesting debugging scenarios
across all mutants --- there are potentially many more but this is how
many we examined. 
Figure~\ref{fig:mutant-breakdown} breaks down the interesting
debugging scenarios produced  for each benchmark by mutator 
but counts are cut off at 50. The mutators produce at least 40 interesting
scenarios for every benchmark, and these scenarios originate from at
least four different mutators per benchmark.  Thus the mutators result in 
a sizeable and diverse population of scenarios for every benchmark.
Furthermore, every mutator contributes scenarios to at least one
benchmark. Especially some mutators  target specific features of
benchmarks, which
justifies their inclusion. For
instance the class-focused mutators are effective in \texttt{take5}
which jives with the fact that \texttt{take5} is the benchmark with 
the most extensive use of object-oriented features.

>>>>

\section{How to Make Lots of Mistakes}

@@

\begin{figure*}
  \begin{tabular}{p{2.1cm} | p{5cm}  | p{5.5cm} }
    {\bf name} & {\bf description} & {\bf example} \\
\hline

@@

The first criterion simply validates that the debugging scenario concerns
an actual type-level mistake. However, The second and third criteria require 
more explanation.  The second one stipulates that the mistake is
detectable under Erasure to ensure that it is detectable by all three
semantics.   Whether a scenario produces a run time error largely depends on 
the underlying semantics but \citet{gfd-oopsla-2019}
show that a run time error under Erasure implies a run time
error under both Transient and Natural. Thus the criterion aligns
interesting debugging scenarios with our lowest denominator, Erasure, 
and excludes  from consideration the very interesting scenarios indeed where Natural or
Transient can detect a mistake but where Erasure is powerless. Of course this choice favors
Erasure over Transient and Natural and, for the same reason, Transient over
Natural. We consider some form of bias towards one or the other semantics
unavoidable no matter how we define interesting debugging scenarios and we
opt for tipping the scales in favor of the theoretically weakest
semantics. We discuss further the implications of this choice in
section~\ref{sec:discussion}.

@@

With the definition of interesting debugging scenarios in hand, we analyze how effective our mutators
are for turning the benchmarks into interesting debugging scenarios.
Indeed the mutators produce at least 72192 interesting debugging scenarios
across all mutants --- there are potentially many more but this is how
many we examined. 
Figure~\ref{fig:mutant-breakdown} breaks down the interesting
debugging scenarios produced  for each benchmark by mutator 
but counts are cut off at 50. The mutators produce at least 40 interesting
scenarios for every benchmark, and these scenarios originate from at
least four different mutators per benchmark.  Thus the mutators result in 
a sizeable and diverse population of scenarios for every benchmark.
Furthermore, every mutator contributes scenarios to at least one
benchmark. Especially some mutators  target specific features of
benchmarks, which
justifies their inclusion. For
instance the class-focused mutators are effective in \texttt{take5}
which jives with the fact that it is the benchmark with 
the most extensive use of object-oriented features.

### 2020-11-19 writing/rational.tex
adding landscape description

\section{How to Simulate the Rational Programmer}

@@

Equipped with these preliminary definitions, we can precisely define what it means
for blame to add value to a gradual type system in a specific setting. 
\begin{quote}
\it
Given a program $\system$, the natural blame mode is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural blame trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural blame trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural blame
trails.\\
\end{quote}

>>>>

\section{How to Test the Rational Programmer}

@@

Equipped with these preliminary definitions, we can precisely define what it means
for blame to add value to a gradual type system for a specific
type-level mistake independently of a program's type annotations. 
\begin{quote}
\it
Given a program $\system$, the natural blame mode is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural blame trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural blame trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural blame
trails.\\
\end{quote}
